---
title: "Implement NLA-GP"
author: "Leonardo Capitani (leocapi07@gmail.com)"
date: "2026-07-19"
date-modified: last-modified
format:
  html:
    css: styles.css
    toc: true
    code-fold: true
    code-tools: true
    code-link: true
    embed-resources: true
    include-after-body:
      text: |
        <script>
          document.addEventListener("DOMContentLoaded", () => {
            document.querySelectorAll("figcaption").forEach(e => {
              e.innerHTML = e.innerHTML.replace(/((.*?)\.)/, "<strong>$1</strong>")
            })
          });
        </script>
    crossref:
      fig-title: "**Figure**"
      tbl-title: "**Table**"
      title-delim: "."

execute:
  echo: true
  warning: false
  message: false

editor: visual
editor_options:
  chunk_output_type: console
---

```{r load_libraries}
# --- Packages used in this notebook -----------------------------------------
# Each library() call loads a set of extra functions into R. Without these,
# the function names used below (fitGP, ggplot, ode, ...) would not exist.
library(GPEDM)     # Gaussian Process EDM: fitGP() (the forecaster) and
                   # makelags() (builds the "time-delay" columns we need)
library(tidyverse) # data wrangling + plotting: dplyr, tidyr, ggplot2, purrr...
library(here)      # builds file paths relative to the project root, so the
                   # code works on any computer without editing paths
library(deSolve)   # numerically solves differential equations: used to
                   # simulate the food-chain model further below
library(zoo)       # helpers for time series (moving averages, interpolation)
```

# Introduction

::: callout-important
**We call this method NLA-GP:** Nested-Library Analysis ([Huang et al. 2024](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1011759)), with [GP-EDM](https://github.com/tanyalrogers/GPEDM) (Munch et al. 2017) instead of the original S-map.
:::

## Background

Regime shifts -- abrupt transitions between alternative ecosystem states -- have been documented across marine, freshwater, and terrestrial systems, often with lasting ecological and economic consequences (Scheffer et al. 2001). Because these transitions are difficult to reverse, correctly identifying when one occurred is central to managing the systems they affect (Folke et al. 2004). Huang et al. 2024 developed Nested-Library Analysis (NLA) to close a specific gap in how such transitions are detected.

::: callout-note
Can we detect when a system's underlying dynamics fundamentally changed -- a true regime shift -- rather than just when its surface statistics (mean, variance, distribution) changed?
:::

Existing change-point methods largely detect shifts in statistical properties of the data, but a generic detection method with theoretical linkage to underlying dynamics has been lacking. This distinction matters because in chaotic systems, a real shift in the governing dynamics can be concealed by the dynamics themselves -- the raw numbers can look statistically similar before and after a shift even though the attractor generating them has actually changed shape.

Since regime shifts can be framed as transitions between alternative attractors (Scheffer & Carpenter 2003), and since Takens' theorem guarantees that an attractor can be reconstructed from time-delayed coordinates of just one observed variable (Takens 1981), the method turns detection into a forecasting question: if we hold out the most recent stretch of the series as a test set and progressively chop data off the far (older) end of the "library" used to reconstruct the attractor, at what cutoff does forecast skill stop improving -- or start improving -- the most?

NLA fixes one end of the time series as a test set to evaluate out-of-sample forecast skill, and repeatedly removes data from the other end, originally using S-map based EDM forecasting.

The logic: if the whole series comes from one stable regime, throwing more historical data into the library should generally help reconstruct the attractor and improve prediction. But if part of that history actually belongs to a different attractor (a prior regime), including it corrupts the reconstruction and forecast skill degrades. So the point where trimming old data maximally improves forecast skill is taken as the change point -- NLA determines the time of a regime shift as the cutting point at which sequential reduction of the library set optimizes the forecast skill of EDM.

How it works:

- Reconstruct the system's dynamics from a time series (via EDM)
- Gradually remove older data ("shrink the library")
- Find the point where prediction accuracy improves the most
- That point is the regime shift: older data comes from a different dynamical regime and actively harms prediction

Huang et al. 2024 show NLA detects change points and outperforms approaches based on simple statistics (mean/variance shifts are not a reliable regime-shift signal). Their original code (S-map based) is [on GitHub](https://github.com/chromian/Nested-Library-Analysis).

## Goal

Replace S-map with GP-EDM inside the NLA framework, per tips from Steve Munch & Tanya Rogers (GP-EDM's authors):

- Fit GP-EDM in **two stages**: hyperparameters are slow to fit, but predicting with them fixed is fast
- To change the training data while keeping hyperparameters fixed: refit with `fitGP()`, passing the previous `pars` as `fixedpars` -- this skips the slow optimization step
- A matrix inversion is still needed whenever the training data changes, but that's cheap

::: callout-warning
**Decision:** we implement NLA-GP using this two-stage approach, and compare it against the original NLA-S-map.
:::

## Algorithm

::: {.callout-tip title="Question"}
Concretely, what does NLA *do* to a time series -- what gets fitted, what gets thrown away, and what number comes out at the end?
:::

Huang et al. 2024 give NLA as two symmetric algorithms ([S1 Text](https://doi.org/10.1371/journal.pcbi.1011759.s001)). The core idea, from their Fig 1 (@fig-nla-concept):

- Fix one end of the time series as a **test set**, used only to check out-of-sample forecast skill -- never to fit anything.
- The rest of the series is the **library**: the data used to reconstruct the system's attractor and forecast the test set.
- Repeatedly chop data off the *far* end of the library (the end away from the test set), so the library gets smaller and smaller.
- If the chopped-off data came from a *different dynamical regime* than the test set (red dots in @fig-nla-concept), removing it should make the forecast *better*. Once only same-regime data is left (grey dots), removing more of it can only make the forecast *worse* -- there's simply less data to fit from.
- The point where this trade-off flips -- error stops improving and starts getting worse -- is the change point.

```{r fig-nla-concept}
#| fig-cap: "The core idea of NLA (redrawn after Huang et al. 2024, Fig 1). Five trials shrink the library by different amounts. Trial 3 removes exactly the misleading (different-regime) data and nothing else -- the best-performing library."
# This chunk draws a *cartoon* of the method. No real data, no model fitting --
# it just colours a grid of dots so the idea is visible at a glance.

true_cp <- 15   # where the (pretend) regime shift happens, in time steps
L       <- 30   # last time step belonging to the library
R       <- 40   # length of the whole (pretend) time series

# One row per "trial". `l` = how many of the oldest points that trial throws
# away. Trial 1 throws away nothing, Trial 5 throws away 30 points.
trials <- tibble(trial = 1:5, l = c(0, 8, 15, 23, 30),
                  label = c("Trial 1", "Trial 2", "Trial 3 (best)", "Trial 4", "Trial 5"))

trials %>%
  # crossing() makes every combination of trial x time step, i.e. one dot per
  # (trial, time) pair -- 5 trials x 40 time steps = 200 dots.
  crossing(time = 1:R) %>%
  # Label each dot. case_when() is a chain of if/else tests, read top to
  # bottom: the FIRST matching condition wins.
  mutate(category = case_when(
    # thrown away AND from before the shift -> misleading data (good riddance)
    time <= pmin(l, true_cp) ~ "removed: different regime",
    # thrown away but from after the shift -> useful data we just lost
    time <= l               ~ "removed: same regime",
    # kept, and used to fit the forecaster
    time <= L                ~ "library (kept)",
    # everything past L is the held-out test set
    TRUE                     ~ "test set"
  )) %>%
  # ggplot builds a plot in layers, combined with "+".
  # x = time, y = the trial label, colour = the category we just assigned.
  # fct_inorder() keeps the trials in the order written above (not alphabetical)
  # and fct_rev() flips them so Trial 1 sits at the top.
  ggplot(aes(time, fct_rev(fct_inorder(label)), color = category)) +
  geom_point(size = 2.5) +                    # draw the dots
  scale_color_manual(values = c(              # pick the colour of each category
    "removed: different regime" = "firebrick",
    "removed: same regime" = "grey70",
    "library (kept)" = "grey20",
    "test set" = "steelblue")) +
  labs(x = "Time", y = NULL, color = NULL) +  # axis titles (NULL = no title)
  theme_minimal() +                           # plain, gridline-only look
  theme(legend.position = "bottom")
```

### Steps

A single shrinking pass (as in @fig-nla-concept) is noisy, so Huang et al. 2024 repeat it from many different **starting points** for the library, then take the median of all the answers. Concretely, for a series of length `R` with the test set fixed at one end and the library ending at `L`:

1. Pick a starting point for the library, call it `n`. For example, with `L = 400` (library ends at row 400): first try `n = 0` (library = rows 1-400), then `n = 50` (library = rows 51-400), then `n = 100`, and so on
2. For that `n`, fit hyperparameters **once**, using the library `{n, ..., L}` to predict the test set
3. Holding those hyperparameters fixed, shrink this *same* library further, from a second point `l` (`l` ranges from `n` up to `L`), recording how the out-of-sample error changes as a curve `E_n(l)` (exactly the "one trial" picture in @fig-nla-concept, but now for library start `n`)
4. If `E_n(.)` is valley-shaped, its minimum is one candidate change point
5. Repeat steps 1-4 for every starting point `n`; the final estimate is the median of all the candidates

::: {.callout-note appearance="minimal"}
In code (below), `n` is this outer library starting point and `l` is how far it's since been trimmed -- matching Huang et al.'s own notation. `nla_left()` always shrinks the **left** library (the early part of the series); `nla_right()` always shrinks the **right** library (the late part) -- see next.
:::

### Left check

Test set = the **last** part of the series. Library shrinks by removing the **oldest** data first (from the start). Finds a shift located in the **early** part of the series.

### Right check

Test set = the **first** part of the series. Library shrinks by removing the **newest** data first (from the end). Finds a shift located in the **late** part of the series.

### Discriminant

A curve is "valley-shaped" exactly when (Huang et al. 2024, Methods, Sec. 2.2):

$$
\frac{d\tilde{\varepsilon}}{dl}(t) < 0 \text{ for } t < \tau, \qquad \frac{d\tilde{\varepsilon}}{dl}(t) > 0 \text{ for } t > \tau
$$

i.e. error strictly decreases while misleading (different-regime) data is still being removed, then strictly increases once only correct (same-regime) data is left to remove. We detect this as a single run of negative differences followed by a single run of positive differences in the smoothed curve.

### Code

::: {.callout-tip title="Question"}
How do those steps translate into R? Which function does which part of the algorithm -- and where does the two-stage speed-up actually happen?
:::

::: {.callout-note appearance="minimal"}
**Reading the code:** every chunk below is commented line by line, so no prior R is assumed. Two idioms recur throughout and are worth knowing up front: `%>%` (the "pipe") passes the result on the left into the function on the right, so `x %>% f()` is just `f(x)`; and `$` pulls one named element out of a list or table, so `fit$pars` means "the `pars` part of `fit`".
:::

We split the implementation into two pieces: (1) fitting GP-EDM and shrinking the library, and (2) smoothing the resulting error curve and applying the discriminant.

::: {.callout-note appearance="minimal"}
**(1) Fitting and shrinking the library** -- `build_lag_frame()`, `fit_hyperparams()`, `refit_fixed()`, `nla_left()`, `nla_right()`:

- `build_lag_frame()`: builds the lag columns once, and z-scores them once, using the *whole series* as a fixed reference (see the scaling callout below)
- `fit_hyperparams()`: the **outer** fit -- for one library starting point `n`, fits GP-EDM hyperparameters from scratch (the slow step)
- `refit_fixed()`: the **inner** refit -- reuses those same hyperparameters via `fixedpars` while the library shrinks further from `l` (only a cheap matrix inversion)
- `nla_left()` / `nla_right()`: loop both of the above over every starting point `n`, exactly as in the Steps above
:::

::: {.callout-warning}
**Scaling, checked once more:** `fitGP()` has its own `scaling` argument (`"global"`/`"local"`) that would z-score the data *for you* -- but recomputed from whatever rows it happens to be given. If we let it do that on every refit, the library's scale would silently drift as it shrinks, while the *fixed* hyperparameters from step 2 stay tuned to the *original* scale -- i.e. the data would effectively get rescaled every refit, on top of hyperparameters that assumed a fixed scale. We avoid this entirely: `build_lag_frame()` z-scores **once**, using the whole series, and every `fitGP()` call below passes `scaling = "none"` so `fitGP()` never rescales again. The data is scaled exactly once, not twice.
:::

```{r core_functions}
#' Build lag/predictor columns ONCE over the full series, and z-score them
#' ONCE using a fixed reference (see the scaling callout above). This is the
#' ONLY place scaling happens -- every fitGP() call below uses
#' scaling = "none" so it is never rescaled again.
build_lag_frame <- function(y, E, tau) {
  # y   = the observed time series, as a plain numeric vector
  # E   = "embedding dimension": how many past values we use as predictors
  # tau = the spacing between those past values (tau = 1 means consecutive)

  T <- length(y)                                   # number of observations
  # A tidy table with one row per time step: a time index and the value.
  base <- data.frame(time = seq_len(T), y = y)

  # makelags() (from GPEDM) adds the "time-delay" predictor columns. With
  # E = 3 and tau = 1 it appends y_1, y_2, y_3 = the value 1, 2 and 3 steps
  # ago. Predicting y from its own recent past is the whole idea behind
  # Empirical Dynamic Modelling: those lagged columns are a stand-in for the
  # (unobserved) state of the system (Takens' theorem).
  lagdf <- makelags(data = base, y = "y", E = E, tau = tau, append = TRUE)

  # Everything that is not "time" or "y" is a predictor column. setdiff()
  # is set subtraction, so xnames ends up as c("y_1", "y_2", "y_3").
  xnames <- setdiff(names(lagdf), c("time", "y"))

  # The first few rows have no "value 3 steps ago", so makelags() puts NA
  # there. complete.cases() flags the rows where nothing is missing; we use
  # only those rows to compute the mean and sd below.
  complete <- stats::complete.cases(lagdf[, c("y", xnames)])

  # z-score the response: subtract the mean, divide by the standard
  # deviation, so the column has mean 0 and sd 1. Done ONCE here, on the
  # whole series, so the scale never changes as the library shrinks.
  ymean <- mean(lagdf$y[complete]); ysd <- sd(lagdf$y[complete])
  lagdf$y <- (lagdf$y - ymean) / ysd

  # Same z-scoring for each predictor column, one at a time.
  # lagdf[[xn]] pulls out the column whose *name* is stored in xn.
  for (xn in xnames) {
    xmean <- mean(lagdf[[xn]][complete]); xsd <- sd(lagdf[[xn]][complete])
    lagdf[[xn]] <- (lagdf[[xn]] - xmean) / xsd
  }

  # R functions return one object, so we bundle both results in a list.
  list(lagdf = lagdf, xnames = xnames)
}

#' Fit GP hyperparameters on library rows [lib_start, lib_end] and evaluate
#' out-of-sample RMSE on test_rows.
fit_hyperparams <- function(lagdf, xnames, lib_start, lib_end, test_rows) {
  # Keep only the rows whose time index falls inside the library window.
  # The "&" means both conditions must hold; the result is the training set.
  train <- lagdf[lagdf$time >= lib_start & lagdf$time <= lib_end, ]

  # tryCatch() runs the code and, if it throws an error, returns NULL instead
  # of stopping the whole notebook. The optimiser inside fitGP() occasionally
  # fails on awkward libraries, and we would rather skip that case than crash.
  tryCatch(
    # fitGP() fits the Gaussian Process forecaster: predict y from the lag
    # columns. scaling = "none" tells it NOT to re-standardise the data (we
    # already did that, once, in build_lag_frame). newdata = test_rows makes
    # it also predict the held-out test set, and report the error there.
    # This is the SLOW call: it searches for the best hyperparameters.
    fitGP(data = train, y = "y", x = xnames, scaling = "none", newdata = test_rows),
    error = function(e) NULL
  )
}

#' Refit with hyperparameters fixed (2-stage trick): only a matrix inversion
#' for the new (smaller) library is recomputed.
refit_fixed <- function(lagdf, xnames, lib_start, lib_end, test_rows, fixedpars) {
  train <- lagdf[lagdf$time >= lib_start & lagdf$time <= lib_end, ]

  # Guard: a fit needs more data points than parameters. na.omit() drops rows
  # with missing values; if too few usable rows remain, give up on this
  # library and return NA (a missing number) rather than a bad fit.
  if (nrow(na.omit(train[, c("y", xnames)])) < length(xnames) + 2) return(NA_real_)

  fit <- tryCatch(
    # Same call as above, but fixedpars hands fitGP() the hyperparameters we
    # already found -- so it SKIPS the slow search and only redoes the cheap
    # linear algebra for this smaller library. This is the two-stage trick.
    fitGP(data = train, y = "y", x = xnames, scaling = "none",
          fixedpars = fixedpars, newdata = test_rows),
    error = function(e) NULL
  )
  if (is.null(fit)) return(NA_real_)  # the fit failed -> report a missing value

  # Pull out the out-of-sample RMSE (root mean squared error on the test set):
  # the single number saying how badly this library forecast the test set.
  # Smaller = better. unname() just strips the "rmse" label off it.
  unname(fit$outsampfitstats["rmse"])
}

#' Left check (Algorithm 1): library = y[1:L], test = y[(L+1):R].
nla_left <- function(y, L, R = length(y), E = 3, tau = 1, Dskipstep = 5,
                      bandwidth = 2 * Dskipstep, min_lib = E * tau + 3) {
  # Arguments (values after "=" are defaults, used if you don't supply them):
  #   L         = last time step of the library; the test set starts at L + 1
  #   R         = last time step used at all (default: end of the series)
  #   Dskipstep = step size of the search -- we try every 5th cut point, not
  #               every single one, to keep the runtime manageable
  #   bandwidth = how much to smooth the error curve before looking for a dip
  #   min_lib   = smallest library we are willing to fit (needs a few more
  #               rows than predictors)

  # Build the lag columns + z-score, once (see build_lag_frame above).
  lf <- build_lag_frame(y, E, tau)
  lagdf <- lf$lagdf; xnames <- lf$xnames

  # The test set: the LAST stretch of the series. It is never trained on.
  test_rows <- lagdf[lagdf$time >= (L + 1) & lagdf$time <= R, ]

  # The outer loop's starting points n: 0, 5, 10, ... Each n is a different
  # "where does the library begin" choice, giving one independent estimate.
  n_seq <- seq(0, L - min_lib, by = Dskipstep)

  # Pre-allocate the results. rep(NA_real_, k) makes a length-k vector of
  # missing numbers, which we fill in as the loop proceeds; entries left as
  # NA mean "no change point found for this n".
  tau_n <- rep(NA_real_, length(n_seq))
  curves <- vector("list", length(n_seq))   # keeps the error curves, for plots

  # ---- Outer loop: one pass per library starting point n -------------------
  for (i in seq_along(n_seq)) {
    n <- n_seq[i]
    lib_start_n <- max(1, n)   # n can be 0, but row indices start at 1

    # STAGE 1 (slow, once per n): fit the hyperparameters on this library.
    hp <- fit_hyperparams(lagdf, xnames, lib_start_n, L, test_rows)
    if (is.null(hp)) next      # "next" = skip to the next n

    # Extract just the hyperparameters we want to hold fixed: one length-scale
    # per predictor, plus two more (process and observation variance).
    fixedpars <- hp$pars[seq_len(length(xnames) + 2)]

    # The inner loop's cut points l: how far the library start has been
    # trimmed forward from n. Each l = one shrunken library.
    l_seq <- seq(n, L - min_lib, by = Dskipstep)
    if (length(l_seq) < 3) next  # too few points to see a valley shape

    # STAGE 2 (fast): for each cut point l, refit with the hyperparameters
    # frozen and record the test-set error. vapply() is a loop that returns
    # one number (numeric(1)) per element of l_seq -- so err is the error
    # curve E_n(l), one value per cut point.
    err <- vapply(l_seq, function(l) {
      refit_fixed(lagdf, xnames, max(1, l), L, test_rows, fixedpars)
    }, numeric(1))

    # Smooth the (noisy) curve, then ask whether it is valley-shaped. If it
    # is, find_valley() returns the position of the dip = the candidate
    # change point for this n; if not, it returns NA.
    y_smooth <- smooth_gaussian(l_seq, err, bandwidth)
    v <- find_valley(l_seq, y_smooth)
    tau_n[i] <- v$tau_hat

    # Store this n's curve so we can plot raw vs. smoothed error later.
    curves[[i]] <- data.frame(n = n, l = l_seq, rmse = err, rmse_smooth = y_smooth)
  }

  # Final answer: the median of all the per-n candidates (robust to the odd
  # wild estimate). If no n produced a candidate at all, return NA.
  list(tau_hat = if (all(is.na(tau_n))) NA_real_ else median(tau_n, na.rm = TRUE),
       tau_n = data.frame(n = n_seq, tau_hat = tau_n),  # the per-n estimates
       # do.call(rbind, curves) stacks the per-n data frames into one table.
       curves = do.call(rbind, curves))
}

#' Right check (Algorithm 2): test = y[1:L], library = y[(L+1):R].
nla_right <- function(y, L, R = length(y), E = 3, tau = 1, Dskipstep = 5,
                       bandwidth = 2 * Dskipstep, min_lib = E * tau + 3) {
  # Mirror image of nla_left(): everything runs backwards in time. The test
  # set is now the EARLY part of the series (rows 1..L), and the library --
  # rows L+1..R -- is trimmed from its late (right-hand) end.
  lf <- build_lag_frame(y, E, tau)
  lagdf <- lf$lagdf; xnames <- lf$xnames

  # Test set = the FIRST stretch of the series this time.
  test_rows <- lagdf[lagdf$time >= 1 & lagdf$time <= L, ]

  # Note the negative step: n counts DOWN from the end of the series.
  n_seq <- seq(R, L + 1 + min_lib, by = -Dskipstep)
  tau_n <- rep(NA_real_, length(n_seq))
  curves <- vector("list", length(n_seq))

  for (i in seq_along(n_seq)) {
    n <- n_seq[i]

    # STAGE 1 (slow): library runs from L+1 up to n -- the start is pinned
    # and the END is what varies, the mirror of the left check.
    hp <- fit_hyperparams(lagdf, xnames, L + 1, n, test_rows)
    if (is.null(hp)) next
    fixedpars <- hp$pars[seq_len(length(xnames) + 2)]

    # Inner cut points, again counting down: trim the library's right edge.
    l_seq <- seq(n, L + 1 + min_lib, by = -Dskipstep)
    if (length(l_seq) < 3) next

    # STAGE 2 (fast): same frozen-hyperparameter refit as in nla_left().
    err <- vapply(l_seq, function(l) {
      refit_fixed(lagdf, xnames, L + 1, l, test_rows, fixedpars)
    }, numeric(1))

    y_smooth <- smooth_gaussian(l_seq, err, bandwidth)
    v <- find_valley(l_seq, y_smooth)
    tau_n[i] <- v$tau_hat
    curves[[i]] <- data.frame(n = n, l = l_seq, rmse = err, rmse_smooth = y_smooth)
  }

  list(tau_hat = if (all(is.na(tau_n))) NA_real_ else median(tau_n, na.rm = TRUE),
       tau_n = data.frame(n = n_seq, tau_hat = tau_n),
       curves = do.call(rbind, curves))
}
```

::: {.callout-note appearance="minimal"}
**(2) Smoothing the error curve and applying the discriminant** -- `smooth_gaussian()`, `find_valley()`:

- `smooth_gaussian()`: the paper's "Gaussian filter" -- a Nadaraya-Watson smoother over the raw `E_n(l)` curve
- `find_valley()`: applies the exact discriminant from the Discriminant section above to the *smoothed* curve, and returns its minimum only if the curve is genuinely valley-shaped (never on a monotone or multi-valley curve)
:::

```{r discriminant_functions}
#' Gaussian-kernel smoothing of an error curve (the paper's "Gaussian filter").
#' bandwidth is in units of the index spacing (Dskipstep).
smooth_gaussian <- function(idx, y, bandwidth) {
  # idx       = positions along the curve (the cut points l)
  # y         = the raw error values at those positions
  # bandwidth = how wide the smoothing window is; bigger = smoother

  # is.finite() is FALSE for NA / NaN / Inf: keep only the usable points.
  keep <- is.finite(y)
  # sum() of a TRUE/FALSE vector counts the TRUEs. With fewer than 3 usable
  # points there is nothing to smooth, so return all-missing.
  if (sum(keep) < 3) return(rep(NA_real_, length(y)))

  out <- rep(NA_real_, length(y))  # container for the smoothed values
  xi <- idx[keep]; yi <- y[keep]   # the usable positions and values

  # For every position, take a weighted average of ALL usable values, where
  # nearby points get a big weight and distant ones almost none. The weights
  # follow a bell curve (hence "Gaussian"). This is a Nadaraya-Watson smoother.
  for (i in seq_along(idx)) {
    w <- exp(-0.5 * ((xi - idx[i]) / bandwidth)^2)  # bell-shaped weights
    out[i] <- sum(w * yi) / sum(w)                  # weighted mean
  }
  out   # the last expression in an R function is what it returns
}

#' Is a (smoothed) error curve valley-shaped, and if so, where is the minimum?
#' Discriminant (Huang et al. 2024, Sec 2.2): the sign sequence of the first
#' difference must be a single run of "-" followed by a single run of "+",
#' with the minimum in the interior.
find_valley <- function(idx, y_smooth, flat_tol = 1e-8) {
  # Drop unusable points, then bail out if too few remain.
  ok <- is.finite(y_smooth)
  if (sum(ok) < 3) return(list(is_valley = FALSE, tau_hat = NA_real_))
  idx <- idx[ok]; y_smooth <- y_smooth[ok]

  # Sort left-to-right by position. (The right check generates its cut points
  # in decreasing order, so this step matters.)
  o <- order(idx); idx <- idx[o]; y_smooth <- y_smooth[o]

  # diff() gives the step-to-step changes: negative = the curve is going
  # down here, positive = going up.
  d <- diff(y_smooth)

  # Treat *tiny* changes as exactly flat (0), so floating-point dust in a
  # genuinely flat stretch is not mistaken for a real up or down move.
  scale_ <- max(abs(y_smooth), flat_tol)
  s <- ifelse(abs(d) < flat_tol * scale_, 0, sign(d))  # -1, 0 or +1
  s_nz <- s[s != 0]                                    # drop the flat steps
  if (length(s_nz) < 2) return(list(is_valley = FALSE, tau_hat = NA_real_))

  # rle() = "run length encoding": it collapses repeats, so
  # c(-1,-1,-1,1,1) becomes the two runs c(-1, 1). A true valley is exactly
  # that: one stretch going down, then one stretch going up. Anything else
  # (still falling, still rising, or several dips) is rejected.
  runs <- rle(s_nz)$values
  is_valley <- length(runs) == 2 && identical(runs, c(-1, 1))

  # which.min() gives the position of the smallest value -- the bottom of the
  # dip. Reject it if it sits at either end of the curve, since that means we
  # never actually saw the error turn back up.
  imin <- which.min(y_smooth)
  at_boundary <- imin == 1 || imin == length(y_smooth)
  is_valley <- is_valley && !at_boundary   # "!" means "not"

  # Report the verdict, and the estimated change point only if it is a valley.
  list(is_valley = is_valley, tau_hat = if (is_valley) idx[imin] else NA_real_)
}
```

::: callout-important
`smooth_gaussian()` and `find_valley()` must be defined before `nla_left()`/`nla_right()` are *called* (they are, in later chunks) but not necessarily before those functions are *defined* -- R only resolves a function's body when it runs, not when it's written. We show them second here because conceptually they come after "shrink the library": you can't smooth or apply the discriminant to a curve until `nla_left()`/`nla_right()` have generated it.
:::

::: {.callout-note appearance="minimal"}
**(3) The original NLA-S-map** -- `smap_predict()`, `fit_smap_theta()`, `fit_hyperparams_smap()`, `refit_fixed_smap()`, `nla_left_smap()`, `nla_right_smap()`. Following Huang et al. 2024's S2 Text: a locally-weighted linear regression around each predictee, with kernel weight `w(d) = exp(-theta * d / dbar)`. `theta` plays the same role as the GP hyperparameters above -- fit once per outer `n` (`fit_hyperparams_smap()`), reused unchanged while the library shrinks further (`refit_fixed_smap()`). `nla_left_smap()`/`nla_right_smap()` mirror `nla_left()`/`nla_right()` exactly, so the two methods are compared apples-to-apples, and reuse the same `smooth_gaussian()`/`find_valley()` from (2).
:::

```{r smap_functions}
#' S-map forecast: for each point we want to predict, fit a straight line
#' through the library, but weight nearby library points much more heavily.
smap_predict <- function(train_x, train_y, test_x, theta, k = NULL) {
  # train_x = library predictors (one row per library point, one column per lag)
  # train_y = the value to be predicted for each library row
  # test_x  = predictors for the points we want to forecast
  # theta   = how sharply to down-weight distant points (0 = plain linear
  #           regression; large = only the closest neighbours matter, i.e.
  #           strongly nonlinear)
  E <- ncol(train_x)                  # number of predictor columns
  if (is.null(k)) k <- E + 1          # how many neighbours define "close"
  N <- nrow(train_x)                  # number of library points

  # Prepend a column of 1s: that column's coefficient becomes the intercept.
  Xtrain1 <- cbind(1, train_x)
  preds <- numeric(nrow(test_x))      # container for the forecasts

  for (i in seq_len(nrow(test_x))) {
    xq <- test_x[i, ]                 # the point we are forecasting ("query")

    # Straight-line (Euclidean) distance from the query to every library
    # point: subtract, square, sum across columns, take the square root.
    d <- sqrt(rowSums((train_x - matrix(xq, N, E, byrow = TRUE))^2))

    # Average distance to the k closest library points -- used as a local
    # yardstick so theta means the same thing in sparse and dense regions.
    dbar <- mean(sort(d)[seq_len(min(k, N))])
    if (dbar <= 0) dbar <- 1e-8       # avoid dividing by zero

    # The S-map kernel: weight falls off exponentially with scaled distance.
    w <- exp(-theta * d / dbar)

    # lm.wfit() = weighted least-squares fit, i.e. a local linear model.
    fit <- tryCatch(lm.wfit(x = Xtrain1, y = train_y, w = w), error = function(e) NULL)

    # Apply that local line to the query point. If the fit failed or produced
    # non-finite coefficients, record a missing value instead.
    preds[i] <- if (is.null(fit) || any(!is.finite(fit$coefficients))) NA_real_ else sum(c(1, xq) * fit$coefficients)
  }
  preds
}

#' Pick the theta that forecasts the test set best -- S-map's one tunable
#' knob, the counterpart of the GP hyperparameters.
fit_smap_theta <- function(train_x, train_y, test_x, test_y) {
  # obj(theta) = the test-set RMSE we get with that theta. Lower is better.
  obj <- function(theta) sqrt(mean((test_y - smap_predict(train_x, train_y, test_x, theta))^2, na.rm = TRUE))
  # optimize() searches the interval for the theta that minimises obj();
  # $minimum extracts the winning theta itself.
  optimize(obj, interval = c(1e-4, 10))$minimum
}

#' The S-map version of fit_hyperparams(): the slow, once-per-n step.
fit_hyperparams_smap <- function(lagdf, xnames, lib_start, lib_end, test_rows) {
  # Slice out the library window and drop rows with missing values.
  train <- na.omit(lagdf[lagdf$time >= lib_start & lagdf$time <= lib_end, c("y", xnames)])
  test <- na.omit(test_rows[, c("y", xnames)])
  # "||" means "or": give up if either set is too small to work with.
  if (nrow(train) < length(xnames) + 2 || nrow(test) == 0) return(NULL)
  # as.matrix() converts the data frame to the plain numeric matrix that
  # smap_predict() expects.
  list(theta = fit_smap_theta(as.matrix(train[, xnames]), train$y, as.matrix(test[, xnames]), test$y))
}

#' The S-map version of refit_fixed(): re-forecast with theta held fixed
#' while the library shrinks, and report the test-set RMSE.
refit_fixed_smap <- function(lagdf, xnames, lib_start, lib_end, test_rows, theta) {
  train <- na.omit(lagdf[lagdf$time >= lib_start & lagdf$time <= lib_end, c("y", xnames)])
  test <- na.omit(test_rows[, c("y", xnames)])
  if (nrow(train) < length(xnames) + 2 || nrow(test) == 0) return(NA_real_)
  p <- smap_predict(as.matrix(train[, xnames]), train$y, as.matrix(test[, xnames]), theta)
  # RMSE: average the squared errors, then take the square root.
  sqrt(mean((test$y - p)^2, na.rm = TRUE))
}

#' Left/right check, with S-map instead of GP-EDM (same nla_left/nla_right
#' structure, so the two methods are compared apples-to-apples).
nla_left_smap <- function(y, L, R = length(y), E = 3, tau = 1, Dskipstep = 5,
                           bandwidth = 2 * Dskipstep, min_lib = E * tau + 3) {
  # Step for step this is nla_left() above; only the forecaster differs, so
  # see there for the detailed walk-through of the loop.
  lf <- build_lag_frame(y, E, tau)
  lagdf <- lf$lagdf; xnames <- lf$xnames
  test_rows <- lagdf[lagdf$time >= (L + 1) & lagdf$time <= R, ]  # test = late part
  n_seq <- seq(0, L - min_lib, by = Dskipstep)                   # library start points
  tau_n <- rep(NA_real_, length(n_seq))

  for (i in seq_along(n_seq)) {
    n <- n_seq[i]
    hp <- fit_hyperparams_smap(lagdf, xnames, max(1, n), L, test_rows)  # fit theta once
    if (is.null(hp)) next
    l_seq <- seq(n, L - min_lib, by = Dskipstep)                        # cut points
    if (length(l_seq) < 3) next
    # Error curve: shrink the library, keeping theta frozen.
    err <- vapply(l_seq, function(l) refit_fixed_smap(lagdf, xnames, max(1, l), L, test_rows, hp$theta), numeric(1))
    # Smooth, then take the bottom of the dip (NA if it isn't valley-shaped).
    v <- find_valley(l_seq, smooth_gaussian(l_seq, err, bandwidth))
    tau_n[i] <- v$tau_hat
  }
  # Median across starting points = the single estimated change point.
  # Median of the per-n candidates; NA if none of them found a valley.
  list(tau_hat = if (all(is.na(tau_n))) NA_real_ else median(tau_n, na.rm = TRUE))
}

nla_right_smap <- function(y, L, R = length(y), E = 3, tau = 1, Dskipstep = 5,
                            bandwidth = 2 * Dskipstep, min_lib = E * tau + 3) {
  # Mirror image of nla_left_smap(), exactly as nla_right() mirrors nla_left():
  # the test set is the early part and the library is trimmed from its late end.
  lf <- build_lag_frame(y, E, tau)
  lagdf <- lf$lagdf; xnames <- lf$xnames
  test_rows <- lagdf[lagdf$time >= 1 & lagdf$time <= L, ]   # test = early part
  n_seq <- seq(R, L + 1 + min_lib, by = -Dskipstep)         # counts down
  tau_n <- rep(NA_real_, length(n_seq))

  for (i in seq_along(n_seq)) {
    n <- n_seq[i]
    hp <- fit_hyperparams_smap(lagdf, xnames, L + 1, n, test_rows)
    if (is.null(hp)) next
    l_seq <- seq(n, L + 1 + min_lib, by = -Dskipstep)
    if (length(l_seq) < 3) next
    err <- vapply(l_seq, function(l) refit_fixed_smap(lagdf, xnames, L + 1, l, test_rows, hp$theta), numeric(1))
    v <- find_valley(l_seq, smooth_gaussian(l_seq, err, bandwidth))
    tau_n[i] <- v$tau_hat
  }
  # Median of the per-n candidates; NA if none of them found a valley.
  list(tau_hat = if (all(is.na(tau_n))) NA_real_ else median(tau_n, na.rm = TRUE))
}
```

# Synthetic test

::: {.callout-tip title="Question"}
Take a system where we know, exactly, when and why a regime shift happens -- and where the shift is *not* obvious in every variable. Does NLA-GP find it -- and does it do so more accurately, or faster, than the original NLA-S-map?
:::

## Simple map

The food-chain system below (Huang et al. 2024's own benchmark) is deliberately hard: a regime shift hidden from most variables by chaos. Before that, a much simpler sanity check: a single variable, no chaos to hide behind, no ODE integration (just an iterated map, so this runs in seconds).

**To answer this, we:**

- Iterate the logistic map `x_{t+1} = r x_t (1 - x_t)`, with `r` set to 2.8 (a stable fixed point) for the first half of the series and to 3.9 (fully chaotic) for the second half
- Add a little observation noise
- Run NLA-GP and NLA-S-map (left + right check) on the raw series

```{r logistic_functions}
#' Simulate the logistic map with an abrupt change in r (fixed point -> chaos).
simulate_logistic <- function(Tmax = 300, cp = 150, r1 = 2.8, r2 = 3.9, sd = 0.01, seed = 1) {
  # Tmax = series length, cp = when the shift happens,
  # r1/r2 = growth rate before/after the shift, sd = observation-noise size.

  # set.seed() makes the random numbers reproducible: same seed, same series.
  set.seed(seed)

  x <- numeric(Tmax)   # empty container of length Tmax (all zeros for now)
  x[1] <- 0.4          # starting value

  # Step the map forward one time unit at a time. Each new value depends only
  # on the previous one -- that is what makes this a "dynamical system".
  for (t in 2:Tmax) {
    r <- if (t < cp) r1 else r2   # the regime shift: r jumps at time cp
    # The logistic map, x_{t+1} = r * x_t * (1 - x_t). The min()/max() pair
    # just clamps the value inside (0, 1) so rounding can't push it out.
    x[t] <- min(max(r * x[t - 1] * (1 - x[t - 1]), 1e-6), 1 - 1e-6)
  }

  # rnorm() draws Tmax random numbers from a bell curve with mean 0: this is
  # the measurement noise, added on top of the true trajectory.
  x + rnorm(Tmax, 0, sd)
}
```

```{r fig-logistic-traj}
#| fig-cap: "Logistic map: a stable fixed point (r=2.8) abruptly becomes chaotic (r=3.9) at t=150 (dashed line)."
y_log <- simulate_logistic()   # simulate with all the defaults above
cp_log <- 150                  # the change point we built in, for reference

# tibble() makes a small table; seq_along() gives 1, 2, ..., length(y_log).
tibble(time = seq_along(y_log), x = y_log) %>%
  ggplot(aes(time, x)) +
  geom_line() +                                     # the trajectory
  # A vertical marker at the true change point, for visual comparison.
  geom_vline(xintercept = cp_log, linetype = "dashed", color = "firebrick") +
  theme_minimal()
```

```{r tbl-logistic-result}
#| tbl-cap: "NLA-GP vs. NLA-S-map on the logistic map: both land close to the true change point (t=150), in a few seconds each."
T_log <- length(y_log)   # how many time steps we have

# Where to split library from test set. For the left check the library covers
# the first 65% of the series (test = the last 35%); for the right check the
# test set is the first 35% (library = the rest).
Ll_log <- round(T_log * 0.65); Lr_log <- round(T_log * 0.35)

# Sys.time() reads the clock: recording it before and after lets us time the
# run, so we can compare how expensive the two methods are.
t0 <- Sys.time()
left_log_gp  <- nla_left(y_log,  L = Ll_log, R = T_log, E = 2, tau = 1, Dskipstep = 6, bandwidth = 12)
right_log_gp <- nla_right(y_log, L = Lr_log, R = T_log, E = 2, tau = 1, Dskipstep = 6, bandwidth = 12)
time_log_gp <- as.numeric(Sys.time() - t0, units = "secs")   # elapsed seconds

# The same two checks, now with the original S-map forecaster.
t0 <- Sys.time()
left_log_sm  <- nla_left_smap(y_log,  L = Ll_log, R = T_log, E = 2, tau = 1, Dskipstep = 6, bandwidth = 12)
right_log_sm <- nla_right_smap(y_log, L = Lr_log, R = T_log, E = 2, tau = 1, Dskipstep = 6, bandwidth = 12)
time_log_sm <- as.numeric(Sys.time() - t0, units = "secs")

# Assemble the results into a table. $tau_hat pulls the estimated change point
# out of each result list; "Combined" is the median of the two checks.
tibble(Method = c("NLA-GP", "NLA-S-map"),
       `Left check` = c(left_log_gp$tau_hat, left_log_sm$tau_hat),
       `Right check` = c(right_log_gp$tau_hat, right_log_sm$tau_hat),
       Combined = c(median(c(left_log_gp$tau_hat, right_log_gp$tau_hat), na.rm = TRUE),
                    median(c(left_log_sm$tau_hat, right_log_sm$tau_hat), na.rm = TRUE)),
       `Runtime (s)` = round(c(time_log_gp, time_log_sm), 1)) %>%
  knitr::kable()   # render it as a formatted table in the HTML output
```

**We conclude:** both methods land within ~15-25 points of the true change point (t=150) in a matter of seconds -- a useful sanity check that the implementation works before moving to a harder, more realistic benchmark.

## Food chain

### Setup

**To answer this, we:**

- Simulate Huang et al. 2024's own validation system (their Sec. 3.1): a 3-species food chain ([Hastings & Powell 1991](https://doi.org/10.2307/1940591)) whose vital rates are coupled to a slowly-forced nutrient resource `N`
- `N`'s forcing crosses a fold bifurcation partway through the series, which abruptly changes the food chain's dynamics -- from chaotic to near-periodic
- Following Huang et al. 2024, run NLA on the variable `y`, which (like their `x`) shows **no obvious shift by eye** despite depending directly on `N`
- Repeat many times (Huang et al. 2024 used 200 replicates) and compare how often NLA-GP and NLA-S-map recover the known change point

::: {.callout-note appearance="minimal"}
**This section simulates the food-chain benchmark:**

- Nutrient resource `N`, forced across a fold bifurcation (as in our earlier synthetic test)
- 3-species food chain `x -> y -> z`, whose mortality/attack-rate parameters depend on `N`
- Adds process noise (random shocks) and measurement noise, as Huang et al. 2024 describe
:::

The model (Huang et al. 2024, @eq-N-forced and @eq-foodchain; parameters in their S3 Text, Table A):

$$
\dot N = a(t) - bN + \frac{cN^m}{N^m+1}, \qquad a(t) = c_4 t + c_5
$$ {#eq-N-forced}

$$
\dot x = x(1-x) - f_1(x)y, \qquad \dot y = -d_1 y + f_1(x)y - f_2(y)z, \qquad \dot z = -d_2 z + f_2(y)z
$$ {#eq-foodchain}

$$
f_i(u) = \frac{p_i u}{1+q_i u}, \qquad d_1 = c_0 N + c_1, \qquad p_1 = c_2 N + c_3
$$ {#eq-fc-params}

@eq-foodchain (with the functional responses and `N`-dependent rates of @eq-fc-params) is exactly the classic Hastings & Powell (1991) chaotic food chain (Huang et al. 2024 cite it directly as their source), reparameterized so that `y`'s mortality (`d_1`) and `x`'s attack rate (`p_1`) both increase with the nutrient resource `N` from @eq-N-forced.

::: callout-warning
**On reproducing Huang et al.'s exact parameters:** `b=0.8, m=8, c=1` (the `N` subsystem), `q1=3.25, p2=0.1, q2=2, d2=0.01`, and `c2=0.05` come directly from their S3 Text parameter table and check out numerically. Two values did not:

- **`d1`'s and `p1`'s baselines appear swapped in the table as extracted.** With `d1` baseline `3.7` and `p1` baseline `0.4`, `p1/q1`'s maximum possible value (~0.13) is always far below `d1`'s minimum, so `y` (and hence `z`) goes extinct from every initial condition we tried, regardless of `N` (verified numerically). Since `x -> y -> z` is explicitly the Hastings & Powell (1991) model, whose well-known chaotic regime uses mortality `m=0.4` and attack rate `a=5`, we swapped the two extracted baselines (`c1=0.4` for `d1`, `c3=3.7` for `p1`) -- this reproduces exactly the expected chaotic food chain.
- **`c0` (how strongly `N` drives up `d1`) needed to be smaller than extracted (`0.75`)** to keep the food chain oscillating (rather than collapsing to extinction) across `N`'s full range; we use `0.15`.

We flag this rather than silently presenting these as the published values.
:::

```{r foodchain_functions}
#' The model's "right-hand side": given the current state, return the rate of
#' change of each variable. deSolve calls this repeatedly to step the system
#' forward in time. (t = current time, state = current values, params = fixed
#' constants.)
foodchain_rhs <- function(t, state, params) {
  # with(as.list(...)) makes every entry of state and params usable by name
  # (N, x, y, z, b, m, ...), so the equations below read like the maths.
  with(as.list(c(state, params)), {
    a  <- c4 * t + c5   # the external forcing on the nutrient: rises with time
    d1 <- c0 * N + c1   # y's death rate -- increases with the nutrient N
    p1 <- c2 * N + c3   # x's attack rate on ... -- also increases with N

    # Nutrient: input a, losses proportional to N, plus a saturating recycling
    # term. That last term is what creates the fold (tipping) bifurcation.
    dN <- a - b * N + c * N^m / (N^m + 1)

    # x = prey: grows logistically, eaten by y.
    # The p1*x*y/(1+q1*x) form is a saturating ("Holling type II") feeding rate.
    dx <- x * (1 - x) - p1 * x * y / (1 + q1 * x)

    # y = mid-level consumer: dies at rate d1, gains from eating x, loses to z.
    dy <- -d1 * y + p1 * x * y / (1 + q1 * x) - p2 * y * z / (1 + q2 * y)

    # z = top predator: dies at rate d2, gains from eating y.
    dz <- -d2 * z + p2 * y * z / (1 + q2 * y)

    # deSolve requires the rates back as a list, in the same order as `state`.
    list(c(N = dN, x = dx, y = dy, z = dz))
  })
}

#' The fixed constants of the model, collected in one place (Huang et al. 2024,
#' S3 Text Table A -- with the two deviations flagged in the callout above).
#' c4 and c5 set the forcing ramp a(t) = c4*t + c5 and are left adjustable.
default_fc_params <- function(c4 = 0.000867, c5 = 0.25) {
  list(b = 0.8, m = 8, c = 1, d2 = 0.01, c0 = 0.15, c1 = 0.4,
       c2 = 0.05, c3 = 3.7, q1 = 3.25, p2 = 0.1, q2 = 2, c4 = c4, c5 = c5)
}

#' Simulate the food chain, with Poisson-arrival process shocks and Gaussian
#' measurement noise (Huang et al. 2024, S3 Text).
simulate_foodchain <- function(Tmax = 1000, dt_sample = 1, lambda = 0.01, rho = 0.05,
                                meas_sd = 0.05, state0 = c(N = 0.3, x = 0.5, y = 0.1, z = 9),
                                params = default_fc_params(), seed = 1) {
  # lambda  = how often random shocks arrive, rho = how big they are
  # meas_sd = size of the measurement noise added at the end
  # state0  = starting abundances of N, x, y, z
  set.seed(seed)   # reproducible randomness

  # --- Process noise: occasional random shocks to the whole system ----------
  # rpois() draws how MANY shocks occur (a Poisson count); runif() draws
  # uniformly random times for them, which we sort into chronological order.
  n_events <- rpois(1, lambda * Tmax)
  event_times <- sort(runif(n_events, 1e-6, Tmax - 1e-6))

  # One multiplier per variable per shock, drawn between 1-rho and 1+rho:
  # e.g. with rho = 0.05, each abundance is nudged by up to +-5%.
  shocks <- matrix(runif(4 * n_events, 1 - rho, 1 + rho), ncol = 4)

  # deSolve calls this at each shock time to modify the state on the spot.
  event_func <- function(t, state, params) {
    i <- which(abs(event_times - t) < 1e-6)[1]   # which shock is this?
    # pmax(..., 1e-8) keeps abundances strictly positive after the nudge.
    if (!is.na(i)) state <- pmax(state * shocks[i, ], 1e-8)
    state
  }

  # --- Solve the differential equations ------------------------------------
  samp_times <- seq(0, Tmax, by = dt_sample)   # times we want output at
  sol <- ode(y = state0, times = samp_times, func = foodchain_rhs, parms = params,
             method = "ode45",   # a standard adaptive-step numerical solver
             # Only pass the shock machinery if there are any shocks at all.
             events = if (n_events > 0) list(func = event_func, time = event_times) else NULL)

  traj <- as.data.frame(sol)               # solver output -> tidy table
  traj <- traj[complete.cases(traj), ]     # drop any rows the solver failed on

  # --- Observation model ---------------------------------------------------
  # z-score each variable (scale()) and add Gaussian measurement noise, so the
  # series we analyse looks like imperfect field data rather than exact maths.
  for (v in c("N", "x", "y", "z")) traj[[v]] <- scale(traj[[v]])[, 1] + rnorm(nrow(traj), 0, meas_sd)

  rownames(traj) <- NULL   # cosmetic: renumber the rows 1, 2, 3, ...
  traj
}

# The forcing a(t) = c4*t + c5 crosses the N subsystem's fold bifurcation
# (at a ~ 0.51) at t = 300, matching Huang et al. 2024's own choice of
# change point (they crop a 1000-step window so this always lands at
# t=300); c4 is derived for this, not taken from the (implausible, as
# extracted: it would sweep a past 1800) supplementary value.
cp_true <- 300
```

```{r fig-foodchain-traj}
#| fig-cap: "One simulated realization (Fig 2-style): z, y, x each depend directly on N, but only z visibly shifts in mean; the shift in x and y is a change in dynamical character (chaotic to near-periodic), not visible by eye. Dashed line: true change point (t=300)."
traj <- simulate_foodchain(seed = 1)   # one realization of the model

traj %>%
  select(time, z, y, x, N) %>%          # keep these columns, in this order
  # pivot_longer() reshapes wide -> long: instead of four value columns we get
  # one "variable" column and one "value" column. ggplot needs this shape to
  # draw one panel per variable. "-time" means "all columns except time".
  pivot_longer(-time, names_to = "variable", values_to = "value") %>%
  # factor(levels = ...) fixes the panel order (top predator first).
  mutate(variable = factor(variable, levels = c("z", "y", "x", "N"))) %>%
  ggplot(aes(time, value)) +
  geom_line() +
  geom_vline(xintercept = cp_true, linetype = "dashed", color = "firebrick") +
  # One stacked panel per variable; free_y lets each use its own y-axis range.
  facet_wrap(~variable, ncol = 1, scales = "free_y") +
  labs(x = "Time", y = NULL) +
  theme_minimal()
```

### GP result

We first demonstrate NLA-GP on a single realization, cropping to a window around the shift for tractability (`t` in `[100, 600]`, so `cp_true` is at crop-index 201).

```{r run_synthetic_gp}
# Analysis settings, reused by every run in this section:
E <- 3            # use 3 lagged values as predictors
tau <- 1          # lags one step apart
Dskipstep <- 15   # try every 15th cut point (coarser = faster)
bandwidth <- 30   # smoothing width for the error curve

# Crop to a window around the shift: the whole 1000-step series would be slow.
crop <- traj[traj$time >= 100 & traj$time <= 600, ]
T_syn <- nrow(crop)          # length of the cropped series
cp_true_crop <- cp_true - 99 # the true change point, renumbered for the crop
y_syn <- crop$y              # the variable we analyse (shows no obvious shift)

# Library/test split, as in the logistic example above.
L_left <- round(T_syn * 0.65); L_right <- round(T_syn * 0.35)

t0 <- Sys.time()   # start the stopwatch
left_gp  <- nla_left(y_syn,  L = L_left, R = T_syn, E = E, tau = tau, Dskipstep = Dskipstep, bandwidth = bandwidth)
right_gp <- nla_right(y_syn, L = L_right, R = T_syn, E = E, tau = tau, Dskipstep = Dskipstep, bandwidth = bandwidth)
time_gp <- as.numeric(Sys.time() - t0, units = "secs")   # elapsed seconds
# Single answer from the two checks: their median (here, just their average).
combined_gp <- median(c(left_gp$tau_hat, right_gp$tau_hat), na.rm = TRUE)

# Side by side with the known truth, so the reader can judge the accuracy.
tibble(estimate = c("True change point", "Left check", "Right check", "Combined"),
       value = c(cp_true_crop, left_gp$tau_hat, right_gp$tau_hat, combined_gp))
```

```{r fig-synthetic-curves}
#| fig-cap: "Right-check NLA-GP error curves for a few outer starting points n. Grey: raw RMSE. Blue: smoothed. Dashed red: true change point."
# Plotting all the outer starting points would be unreadable, so pick 4 of
# them, evenly spaced: list the unique n values, sort them, and take the
# 1st, ~1/3, ~2/3 and last. ("." is the value flowing through the pipe.)
n_show <- unique(right_gp$curves$n) %>% sort() %>% { .[round(seq(1, length(.), length.out = 4))] }

right_gp$curves %>%
  filter(n %in% n_show) %>%      # keep only those 4 starting points
  ggplot(aes(l, rmse)) +
  geom_point(color = "grey60", size = 1) +                                  # raw error
  geom_line(aes(y = rmse_smooth), color = "steelblue", linewidth = 0.8) +   # smoothed
  geom_vline(xintercept = cp_true_crop, linetype = "dashed", color = "firebrick") +
  facet_wrap(~n, labeller = label_both) +
  labs(x = "Library start l", y = "Out-of-sample RMSE") +
  theme_minimal()
```

### Smap result

We use the original NLA-S-map, defined in the Algorithm/Code section above, on the same cropped window.

```{r run_synthetic_smap}
# Exactly the same inputs and settings as the GP run above -- only the
# forecaster differs -- so the comparison is like for like.
t0 <- Sys.time()
left_sm  <- nla_left_smap(y_syn,  L = L_left, R = T_syn, E = E, tau = tau, Dskipstep = Dskipstep, bandwidth = bandwidth)
right_sm <- nla_right_smap(y_syn, L = L_right, R = T_syn, E = E, tau = tau, Dskipstep = Dskipstep, bandwidth = bandwidth)
time_sm <- as.numeric(Sys.time() - t0, units = "secs")
combined_sm <- median(c(left_sm$tau_hat, right_sm$tau_hat), na.rm = TRUE)

tibble(estimate = c("True change point", "Left check", "Right check", "Combined"),
       value = c(cp_true_crop, left_sm$tau_hat, right_sm$tau_hat, combined_sm))
```

```{r tbl-synthetic-single}
#| tbl-cap: "NLA-GP vs. NLA-S-map on this single realization: accuracy and runtime."
tibble(Method = c("NLA-GP", "NLA-S-map"),
       `Left check` = c(left_gp$tau_hat, left_sm$tau_hat),
       `Right check` = c(right_gp$tau_hat, right_sm$tau_hat),
       Combined = c(combined_gp, combined_sm),
       # abs() = distance from the truth regardless of direction (early or late).
       `Error (vs. true)` = round(abs(c(combined_gp, combined_sm) - cp_true_crop)),
       `Runtime (s)` = round(c(time_gp, time_sm), 1)) %>%
  knitr::kable()
```

### Comparison

A single realization isn't a reliable performance measure (Huang et al. 2024 use 200 replicates, Fig 4). We repeat the simulation with different noise draws and score both methods over many replicates.

::: callout-note
**Compute:** we use 20 replicates, not 200 -- each replicate takes ~25-40s for both methods combined, so 200 would take over 2 hours. This is far fewer than Huang et al. 2024 used, so treat the resulting distributions as indicative, not a precise reproduction of their Fig 4.
:::

```{r run_replicates}
N_REPLICATES <- 20   # how many independent simulated datasets to score

#' Everything done above for one dataset, wrapped up so we can repeat it. The
#' seed both identifies the replicate and gives it its own noise draw.
run_one_replicate <- function(seed) {
  tr <- simulate_foodchain(seed = seed)     # a fresh simulated dataset
  cr <- tr[tr$time >= 100 & tr$time <= 600, ]  # same crop window as before
  y <- cr$y                                 # same variable
  Tn <- nrow(cr)
  Ll <- round(Tn * 0.65); Lr <- round(Tn * 0.35)   # same library/test split

  # All four runs: GP and S-map, left check and right check.
  lg <- nla_left(y,  L = Ll, R = Tn, E = E, tau = tau, Dskipstep = Dskipstep, bandwidth = bandwidth)
  rg <- nla_right(y, L = Lr, R = Tn, E = E, tau = tau, Dskipstep = Dskipstep, bandwidth = bandwidth)
  ls <- nla_left_smap(y,  L = Ll, R = Tn, E = E, tau = tau, Dskipstep = Dskipstep, bandwidth = bandwidth)
  rs <- nla_right_smap(y, L = Lr, R = Tn, E = E, tau = tau, Dskipstep = Dskipstep, bandwidth = bandwidth)

  # One row of results for this replicate: each method's combined estimate.
  tibble(seed = seed,
         `NLA-GP` = median(c(lg$tau_hat, rg$tau_hat), na.rm = TRUE),
         `NLA-S-map` = median(c(ls$tau_hat, rs$tau_hat), na.rm = TRUE))
}

# map_dfr() runs the function for seeds 1..N and stacks the one-row results
# into a single data frame. This is the slow chunk: expect several minutes.
replicate_results <- map_dfr(seq_len(N_REPLICATES), run_one_replicate)

# Reshape to one row per (replicate, method), which is what the summary table
# and the plots below expect.
replicate_long <- replicate_results %>%
  pivot_longer(-seed, names_to = "Method", values_to = "tau_hat")
```

```{r tbl-replicate-summary}
#| tbl-cap: "NLA-GP vs. NLA-S-map across replicate simulations: detection rate, median estimate, and error relative to the true change point (crop-index 201)."
replicate_long %>%
  group_by(Method) %>%   # compute the following separately for each method
  summarise(
    # !is.na(tau_hat) is TRUE when a change point WAS found. Summing TRUEs
    # counts them; averaging them gives the proportion.
    `Detected (n)` = sum(!is.na(tau_hat)),
    `Detected (%)` = round(100 * mean(!is.na(tau_hat))),
    # Typical estimate, and typical distance from the truth. We use medians
    # (not means) so one wild replicate can't dominate the summary.
    `Median estimate` = median(tau_hat, na.rm = TRUE),
    `Median error` = median(abs(tau_hat - cp_true_crop), na.rm = TRUE),
    .groups = "drop") %>%   # ungroup afterwards, to avoid surprises later
  knitr::kable()
```

```{r fig-replicate-density}
#| fig-cap: "Distribution of detected change points across replicates (Fig 4-style, left panel). Dashed red: true change point."
# filter(!is.na(...)) keeps only the replicates where a change point was found.
ggplot(replicate_long %>% filter(!is.na(tau_hat)), aes(tau_hat, fill = Method)) +
  # Histogram: bars count how many replicates landed in each range.
  # position = "identity" + alpha = 0.5 overlays the two methods semi-transparently.
  geom_histogram(position = "identity", alpha = 0.5, bins = 15) +
  geom_vline(xintercept = cp_true_crop, linetype = "dashed", color = "firebrick") +
  labs(x = "Detected change point", y = "Count") +
  theme_minimal()
```

```{r fig-replicate-ecdf}
#| fig-cap: "Empirical cumulative distribution of detected change points (Fig 4-style, right panel). Dashed red: true change point."
ggplot(replicate_long %>% filter(!is.na(tau_hat)), aes(tau_hat, color = Method)) +
  # stat_ecdf() draws the empirical cumulative distribution: at each x it shows
  # the fraction of replicates that estimated a change point at or before x.
  # A curve that jumps steeply near the true value = tightly clustered estimates.
  stat_ecdf(linewidth = 0.8) +
  geom_vline(xintercept = cp_true_crop, linetype = "dashed", color = "firebrick") +
  labs(x = "Detected change point", y = "Cumulative count (fraction)") +
  theme_minimal()
```

**We conclude:**

- Both methods detect *a* change point in most replicates (NLA-S-map: 20/20; NLA-GP: 19/20, @tbl-replicate-summary), but **NLA-S-map is substantially more accurate**: its median error is 11 (index units), against 109 for NLA-GP -- an order of magnitude apart, not noise
- @fig-replicate-density and @fig-replicate-ecdf show why: NLA-S-map's estimates cluster tightly around the true change point, while NLA-GP's are detected almost as often but scattered far more widely
- This is the opposite of what we expected going in (that GP-EDM's flexibility should help on a genuinely chaotic system). One plausible reason: this food chain's attractor is thin and fast-changing, so a squared-exponential GP kernel fit from a modest library may struggle to track it as precisely as S-map's simpler, more local linear approximation -- worth checking with a larger `E` or more library data before concluding GP-EDM is worse here in general
- Takeaway: **on this benchmark, NLA-GP does not outperform the original NLA-S-map** -- if anything, the reverse. Any claim that GP-EDM improves NLA needs to be demonstrated on a system where it actually wins, not assumed

# Conclusion

- Implemented NLA-GP: Huang et al. 2024's NLA algorithm with GP-EDM instead of S-map (two-stage fitting, per Munch & Rogers)
- Validated on Huang et al. 2024's own food-chain benchmark (a known regime shift, hidden from `x` and `y` by eye): both methods detect a change point in nearly all replicates, but NLA-S-map is markedly more accurate (median error 11 vs. 109) -- NLA-GP did not outperform the original method here
- See `2_empirical.qmd` for NLA-GP applied to real data (Lake Zurich plankton, all 13 functional groups) and compared against Medeiros et al. 2025's independent change-point analysis

## Next

- Investigate *why* NLA-GP underperformed NLA-S-map here: try larger `E`, more library data per fit, or a longer (less compute-constrained) library/test split
- Sensitivity analysis over `E`, `tau`, `Dskipstep`, and the smoothing bandwidth (we used one reasonable default throughout, not a tuned choice)
- Run the full 200-replicate comparison (compute permitting) for a direct comparison to Huang et al. 2024's Fig 4
- Apply NLA-GP to other empirical regime-shift datasets (e.g. Pacific Decadal Oscillation, per Huang et al. 2024 Sec. 3.2)
