---
title: "NLA-GP: Empirical Test"
author: "Leonardo Capitani (leocapi07@gmail.com)"
date: "2026-07-19"
date-modified: last-modified
format:
  html:
    css: styles.css
    toc: true
    code-fold: true
    code-tools: true
    code-link: true
    embed-resources: true
    crossref:
      fig-title: "**Figure**"
      tbl-title: "**Table**"
      title-delim: "."

execute:
  echo: true
  warning: false
  message: false

editor: visual
editor_options:
  chunk_output_type: console
---

```{r load_libraries}
# --- Packages used in this notebook -----------------------------------------
library(GPEDM)     # Gaussian Process EDM: fitGP() and makelags()
library(tidyverse) # data wrangling + plotting (dplyr, tidyr, ggplot2, purrr)
library(here)      # file paths relative to the project root
library(zoo)       # time-series helpers: na.approx() and rollmean() below
```

::: callout-important
This notebook applies **NLA-GP** to real data: Lake Zurich plankton (Medeiros et al. 2025) and the Pacific Decadal Oscillation index (Huang et al. 2024, Sec. 3.2). See `1_implement.qmd` for the algorithm, its derivation, and validation on synthetic systems with a known regime shift. The NLA-GP (and NLA-S-map) implementation is reproduced below so this notebook runs standalone.
:::

## Code

::: {.callout-tip title="Question"}
What exactly is being re-run here, and why repeat it instead of loading it from `1_implement.qmd`?
:::

::: {.callout-note appearance="minimal"}
**Answer:** this is the same NLA-GP and NLA-S-map implementation, copied verbatim so this notebook renders standalone -- nothing new is introduced. The comments below are a condensed reminder; `1_implement.qmd` has the full line-by-line walk-through and the reasoning behind the design (especially why the data is z-scored exactly once).
:::

```{r core_functions}
# NOTE: every function in this chunk is copied verbatim from `1_implement.qmd`,
# so this notebook runs on its own. See that notebook for the full, line-by-line
# walk-through; the comments here are a condensed reminder of what each does.

#' Gaussian-kernel smoothing of an error curve (the paper's "Gaussian filter").
smooth_gaussian <- function(idx, y, bandwidth) {
  keep <- is.finite(y)   # usable points only (no NA / NaN / Inf)
  if (sum(keep) < 3) return(rep(NA_real_, length(y)))   # too little to smooth
  out <- rep(NA_real_, length(y))
  xi <- idx[keep]; yi <- y[keep]
  # At each position, take a weighted average of all values, with bell-shaped
  # weights so nearby points count most and distant ones barely at all.
  for (i in seq_along(idx)) {
    w <- exp(-0.5 * ((xi - idx[i]) / bandwidth)^2)
    out[i] <- sum(w * yi) / sum(w)
  }
  out
}

#' Is a (smoothed) error curve valley-shaped, and if so, where is the minimum?
#' Discriminant (Huang et al. 2024, Sec 2.2): sign of the first difference is
#' a single run of "-" followed by a single run of "+".
find_valley <- function(idx, y_smooth, flat_tol = 1e-8) {
  ok <- is.finite(y_smooth)
  if (sum(ok) < 3) return(list(is_valley = FALSE, tau_hat = NA_real_))
  idx <- idx[ok]; y_smooth <- y_smooth[ok]
  o <- order(idx); idx <- idx[o]; y_smooth <- y_smooth[o]   # sort left to right

  d <- diff(y_smooth)   # step-to-step changes: negative = falling, positive = rising
  # Treat negligible wiggles as exactly flat, so numerical dust isn't read as a move.
  scale_ <- max(abs(y_smooth), flat_tol)
  s <- ifelse(abs(d) < flat_tol * scale_, 0, sign(d))
  s_nz <- s[s != 0]
  if (length(s_nz) < 2) return(list(is_valley = FALSE, tau_hat = NA_real_))

  # rle() collapses repeats: a true valley is exactly one falling run followed
  # by one rising run, i.e. c(-1, 1). Anything else is rejected.
  runs <- rle(s_nz)$values
  is_valley <- length(runs) == 2 && identical(runs, c(-1, 1))

  imin <- which.min(y_smooth)   # position of the bottom of the dip
  # Reject a minimum sitting at either end: we never saw the error turn back up.
  at_boundary <- imin == 1 || imin == length(y_smooth)
  is_valley <- is_valley && !at_boundary

  list(is_valley = is_valley, tau_hat = if (is_valley) idx[imin] else NA_real_)
}

#' Build lag/predictor columns ONCE over the full series, and z-score them
#' ONCE using a fixed reference (see 1_implement.qmd for why).
build_lag_frame <- function(y, E, tau) {
  T <- length(y)
  base <- data.frame(time = seq_len(T), y = y)
  # makelags() adds the "value E steps ago" predictor columns -- the time-delay
  # embedding that stands in for the system's unobserved state.
  lagdf <- makelags(data = base, y = "y", E = E, tau = tau, append = TRUE)
  xnames <- setdiff(names(lagdf), c("time", "y"))   # the predictor column names

  # z-score everything ONCE, using only the rows with no missing values, so the
  # scale stays fixed as the library shrinks later on.
  complete <- stats::complete.cases(lagdf[, c("y", xnames)])
  ymean <- mean(lagdf$y[complete]); ysd <- sd(lagdf$y[complete])
  lagdf$y <- (lagdf$y - ymean) / ysd
  for (xn in xnames) {
    xmean <- mean(lagdf[[xn]][complete]); xsd <- sd(lagdf[[xn]][complete])
    lagdf[[xn]] <- (lagdf[[xn]] - xmean) / xsd
  }
  list(lagdf = lagdf, xnames = xnames)
}

#' Stage 1 (slow): fit the GP hyperparameters on this library.
fit_hyperparams <- function(lagdf, xnames, lib_start, lib_end, test_rows) {
  train <- lagdf[lagdf$time >= lib_start & lagdf$time <= lib_end, ]   # library rows
  # tryCatch returns NULL instead of crashing if the optimiser fails.
  tryCatch(
    # scaling = "none": we already standardised, in build_lag_frame().
    fitGP(data = train, y = "y", x = xnames, scaling = "none", newdata = test_rows),
    error = function(e) NULL
  )
}

#' Stage 2 (fast): refit on a smaller library with those hyperparameters frozen,
#' and return the test-set RMSE (lower = better forecast).
refit_fixed <- function(lagdf, xnames, lib_start, lib_end, test_rows, fixedpars) {
  train <- lagdf[lagdf$time >= lib_start & lagdf$time <= lib_end, ]
  # Not enough usable rows to fit? Report a missing value rather than a bad fit.
  if (nrow(na.omit(train[, c("y", xnames)])) < length(xnames) + 2) return(NA_real_)
  fit <- tryCatch(
    # fixedpars skips the slow hyperparameter search entirely.
    fitGP(data = train, y = "y", x = xnames, scaling = "none",
          fixedpars = fixedpars, newdata = test_rows),
    error = function(e) NULL
  )
  if (is.null(fit)) return(NA_real_)
  unname(fit$outsampfitstats["rmse"])   # out-of-sample error, label stripped
}

#' Left check (Algorithm 1): library = y[1:L], test = y[(L+1):R].
nla_left <- function(y, L, R = length(y), E = 3, tau = 1, Dskipstep = 5,
                      bandwidth = 2 * Dskipstep, min_lib = E * tau + 3) {
  # L = last library time step; Dskipstep = search every Dskipstep-th cut point;
  # bandwidth = smoothing width; min_lib = smallest library worth fitting.
  lf <- build_lag_frame(y, E, tau)
  lagdf <- lf$lagdf; xnames <- lf$xnames
  test_rows <- lagdf[lagdf$time >= (L + 1) & lagdf$time <= R, ]   # test = late part

  n_seq <- seq(0, L - min_lib, by = Dskipstep)   # outer library starting points
  tau_n <- rep(NA_real_, length(n_seq))          # one candidate change point per n
  curves <- vector("list", length(n_seq))        # error curves, kept for plotting

  # ---- One pass per starting point n --------------------------------------
  for (i in seq_along(n_seq)) {
    n <- n_seq[i]
    lib_start_n <- max(1, n)   # n may be 0, but row numbers start at 1

    hp <- fit_hyperparams(lagdf, xnames, lib_start_n, L, test_rows)  # slow, once
    if (is.null(hp)) next      # the fit failed -> skip to the next n
    # The hyperparameters to freeze: one per predictor, plus two variance terms.
    fixedpars <- hp$pars[seq_len(length(xnames) + 2)]

    l_seq <- seq(n, L - min_lib, by = Dskipstep)   # inner cut points
    if (length(l_seq) < 3) next                    # too short to show a valley

    # The error curve: forecast skill as the library is trimmed step by step.
    err <- vapply(l_seq, function(l) {
      refit_fixed(lagdf, xnames, max(1, l), L, test_rows, fixedpars)
    }, numeric(1))

    # Smooth it, then take the bottom of the dip -- or NA if it's not a valley.
    y_smooth <- smooth_gaussian(l_seq, err, bandwidth)
    v <- find_valley(l_seq, y_smooth)
    tau_n[i] <- v$tau_hat
    curves[[i]] <- data.frame(n = n, l = l_seq, rmse = err, rmse_smooth = y_smooth)
  }

  # Final answer = median of the per-n candidates (robust to the odd wild one);
  # NA if no starting point produced a valley at all. We also return the raw
  # per-n estimates and all the error curves (stacked with rbind) for plotting.
  list(tau_hat = if (all(is.na(tau_n))) NA_real_ else median(tau_n, na.rm = TRUE),
       tau_n = data.frame(n = n_seq, tau_hat = tau_n),
       curves = do.call(rbind, curves))
}

#' Right check (Algorithm 2): test = y[1:L], library = y[(L+1):R].
nla_right <- function(y, L, R = length(y), E = 3, tau = 1, Dskipstep = 5,
                       bandwidth = 2 * Dskipstep, min_lib = E * tau + 3) {
  # Mirror image of nla_left(): the test set is now the EARLY part of the
  # series, and the library (rows L+1..R) is trimmed from its LATE end.
  lf <- build_lag_frame(y, E, tau)
  lagdf <- lf$lagdf; xnames <- lf$xnames
  test_rows <- lagdf[lagdf$time >= 1 & lagdf$time <= L, ]   # test = early part

  n_seq <- seq(R, L + 1 + min_lib, by = -Dskipstep)   # note: counts DOWN
  tau_n <- rep(NA_real_, length(n_seq))
  curves <- vector("list", length(n_seq))

  for (i in seq_along(n_seq)) {
    n <- n_seq[i]
    # Library start is pinned at L+1; its END (n) is what varies.
    hp <- fit_hyperparams(lagdf, xnames, L + 1, n, test_rows)
    if (is.null(hp)) next
    fixedpars <- hp$pars[seq_len(length(xnames) + 2)]

    l_seq <- seq(n, L + 1 + min_lib, by = -Dskipstep)   # trim the right edge
    if (length(l_seq) < 3) next
    err <- vapply(l_seq, function(l) {
      refit_fixed(lagdf, xnames, L + 1, l, test_rows, fixedpars)
    }, numeric(1))

    y_smooth <- smooth_gaussian(l_seq, err, bandwidth)
    v <- find_valley(l_seq, y_smooth)
    tau_n[i] <- v$tau_hat
    curves[[i]] <- data.frame(n = n, l = l_seq, rmse = err, rmse_smooth = y_smooth)
  }

  # Final answer = median of the per-n candidates (robust to the odd wild one);
  # NA if no starting point produced a valley at all. We also return the raw
  # per-n estimates and all the error curves (stacked with rbind) for plotting.
  list(tau_hat = if (all(is.na(tau_n))) NA_real_ else median(tau_n, na.rm = TRUE),
       tau_n = data.frame(n = n_seq, tau_hat = tau_n),
       curves = do.call(rbind, curves))
}
```

The original NLA-S-map (Huang et al. 2024's S2 Text) is reproduced too, so we can compare it against NLA-GP on the PDO test below, as we did on synthetic data in `1_implement.qmd`.

```{r smap_functions}
#' S-map forecast: predict each test point with a straight line fitted through
#' the library, weighting nearby library points far more heavily than distant
#' ones. theta controls how sharp that down-weighting is (0 = plain linear
#' regression; large = strongly local, i.e. nonlinear).
smap_predict <- function(train_x, train_y, test_x, theta, k = NULL) {
  E <- ncol(train_x)             # number of predictor columns
  if (is.null(k)) k <- E + 1     # how many neighbours count as "close"
  N <- nrow(train_x)             # number of library points
  Xtrain1 <- cbind(1, train_x)   # leading 1s column -> gives us an intercept
  preds <- numeric(nrow(test_x))

  for (i in seq_len(nrow(test_x))) {
    xq <- test_x[i, ]            # the point being forecast
    # Euclidean distance from xq to every library point.
    d <- sqrt(rowSums((train_x - matrix(xq, N, E, byrow = TRUE))^2))
    # Local yardstick: mean distance to the k nearest library points.
    dbar <- mean(sort(d)[seq_len(min(k, N))])
    if (dbar <= 0) dbar <- 1e-8  # never divide by zero
    w <- exp(-theta * d / dbar)  # the S-map kernel weights
    # Weighted least squares = a local linear model around this query point.
    fit <- tryCatch(lm.wfit(x = Xtrain1, y = train_y, w = w), error = function(e) NULL)
    preds[i] <- if (is.null(fit) || any(!is.finite(fit$coefficients))) NA_real_ else sum(c(1, xq) * fit$coefficients)
  }
  preds
}

#' Pick the theta with the best test-set forecast: S-map's single tunable knob,
#' the counterpart of the GP hyperparameters.
fit_smap_theta <- function(train_x, train_y, test_x, test_y) {
  obj <- function(theta) sqrt(mean((test_y - smap_predict(train_x, train_y, test_x, theta))^2, na.rm = TRUE))
  optimize(obj, interval = c(1e-4, 10))$minimum   # the theta minimising the error
}

#' Stage 1 for S-map (slow, once per starting point): estimate theta.
fit_hyperparams_smap <- function(lagdf, xnames, lib_start, lib_end, test_rows) {
  train <- na.omit(lagdf[lagdf$time >= lib_start & lagdf$time <= lib_end, c("y", xnames)])
  test <- na.omit(test_rows[, c("y", xnames)])
  if (nrow(train) < length(xnames) + 2 || nrow(test) == 0) return(NULL)   # too little data
  list(theta = fit_smap_theta(as.matrix(train[, xnames]), train$y, as.matrix(test[, xnames]), test$y))
}

#' Stage 2 for S-map (fast): re-forecast on a smaller library with theta frozen,
#' and report the test-set RMSE.
refit_fixed_smap <- function(lagdf, xnames, lib_start, lib_end, test_rows, theta) {
  train <- na.omit(lagdf[lagdf$time >= lib_start & lagdf$time <= lib_end, c("y", xnames)])
  test <- na.omit(test_rows[, c("y", xnames)])
  if (nrow(train) < length(xnames) + 2 || nrow(test) == 0) return(NA_real_)
  p <- smap_predict(as.matrix(train[, xnames]), train$y, as.matrix(test[, xnames]), theta)
  sqrt(mean((test$y - p)^2, na.rm = TRUE))   # root mean squared error
}

#' Right check with S-map instead of GP-EDM -- structurally identical to
#' nla_right() above, so the two methods can be compared like for like.
nla_right_smap <- function(y, L, R = length(y), E = 3, tau = 1, Dskipstep = 5,
                            bandwidth = 2 * Dskipstep, min_lib = E * tau + 3) {
  lf <- build_lag_frame(y, E, tau)
  lagdf <- lf$lagdf; xnames <- lf$xnames
  test_rows <- lagdf[lagdf$time >= 1 & lagdf$time <= L, ]   # test = early part
  n_seq <- seq(R, L + 1 + min_lib, by = -Dskipstep)         # counts down
  tau_n <- rep(NA_real_, length(n_seq))

  for (i in seq_along(n_seq)) {
    n <- n_seq[i]
    hp <- fit_hyperparams_smap(lagdf, xnames, L + 1, n, test_rows)   # fit theta once
    if (is.null(hp)) next
    l_seq <- seq(n, L + 1 + min_lib, by = -Dskipstep)                # cut points
    if (length(l_seq) < 3) next
    # Error curve with theta held fixed while the library shrinks.
    err <- vapply(l_seq, function(l) refit_fixed_smap(lagdf, xnames, L + 1, l, test_rows, hp$theta), numeric(1))
    v <- find_valley(l_seq, smooth_gaussian(l_seq, err, bandwidth))  # smooth, then find the dip
    tau_n[i] <- v$tau_hat
  }
  # Median of the per-n candidates; NA if none of them found a valley.
  list(tau_hat = if (all(is.na(tau_n))) NA_real_ else median(tau_n, na.rm = TRUE))
}
```

# Lake Zurich

::: {.callout-tip title="Question"}
Lake Zurich shifted from a eutrophic to an oligotrophic state in the late 1980s. [Medeiros et al. 2025](https://www.pnas.org/doi/10.1073/pnas.2416637122) dated this shift separately for all 13 plankton functional groups (@fig-medeiros-s17, @tbl-medeiros-s1). Does NLA-GP find similar timing, on the full monthly time series from Lake Zurich ([Merz et al. 2023](https://www.nature.com/articles/s41558-023-01615-6))?
:::

## Data

**To answer this, we:**

- Process the raw monthly counts: interpolate a handful of missing months, z-score (`scale()`) each variable
- Use the **full** 1977-2020 series (no date restriction)
- Run NLA-GP (left + right check) on **all 13** functional groups

::: {.callout-note appearance="minimal"}
**This section loads and processes the Lake Zurich data:**

- Builds a monthly date index, interpolates a few missing months
- Renames columns, z-scores each variable (`scale()`, no log-transform)
- `zurich_month_index()`: year/month -> row index, for marking reference dates
:::

```{r zurich_functions}
zurich_long_names <- c(
  Cy = "Cyanobacteria", Gr1 = "Green algae (small)", Gr2 = "Green algae (large)",
  Di1 = "Diatoms (small)", Di2 = "Diatoms (large)", Go1 = "Gold algae (small)",
  Go2 = "Gold algae (large)", Cr1 = "Cryptophytes (small)", Cr2 = "Cryptophytes (large)",
  Mi = "Mixotrophic flagellates", He = "Large herbivores", Om = "Omnivores",
  Pr = "Invertebrate predators", temperature = "Temperature", phosphorus = "Phosphate"
)

#' Load and process the raw Lake Zurich CSV (all available months): builds a
#' monthly time index, linearly interpolates the handful of scattered
#' single-month gaps, and z-scores each variable.
process_zurich_data <- function(path) {
  z <- read.csv(path)                 # read the raw comma-separated file
  z <- z[order(z$year, z$month), ]    # sort into chronological order
  z$time_idx <- seq_len(nrow(z))      # 1, 2, 3, ... = position in the series
  # sprintf() glues year and month into "1977-04-01", which as.Date() turns
  # into a real date R can plot on a time axis. %02d pads months to two digits.
  z$date <- as.Date(sprintf("%d-%02d-01", z$year, z$month))

  # The 13 plankton groups = everything named except the two environmental
  # variables (setdiff() is set subtraction).
  plankton_vars <- setdiff(names(zurich_long_names), c("temperature", "phosphorus"))

  # na.approx() fills gaps by drawing a straight line between the observations
  # on either side. Only a handful of single months are missing.
  for (v in c(plankton_vars, "temperature", "phosphorus")) {
    z[[v]] <- na.approx(z[[v]], x = z$time_idx, na.rm = FALSE)
  }
  # Drop any rows still incomplete (gaps at the very start or end can't be
  # interpolated, since there is nothing on one side to interpolate from).
  z <- z[complete.cases(z[, plankton_vars]), ]

  # z-score each group into a new "scaled_..." column, so all 13 are on a
  # comparable scale (mean 0, sd 1) regardless of how abundant they are.
  for (v in plankton_vars) {
    z[[paste0("scaled_", v)]] <- as.numeric(scale(z[[v]]))
  }

  z$time_idx <- seq_len(nrow(z))   # renumber, since rows may have been dropped
  rownames(z) <- NULL
  z
}

#' Row index closest to a given year-month, for marking reference dates.
zurich_month_index <- function(z, year, month) {
  target <- as.Date(sprintf("%d-%02d-01", year, month))
  # which.min(abs(...)) = the row whose date is nearest to the target date.
  which.min(abs(z$date - target))
}

# here() builds the path from the project root, so this works on any machine.
z <- process_zurich_data(here("data", "raw", "lake_zurich.csv"))
```

![This figure was made by Medeiros et al. 2025 (their Fig. S17). Monthly time series for all 13 plankton functional groups (in ind/l), phosphate concentration (in mg/l), and water temperature (in Celsius) from Lake Zurich between January 1978 and December 2019. As explained in the Materials and Methods section in the main text, Medeiros et al. 2025 performed a change-point analysis to determine the location of a potential tipping point for each functional group independently. This potential tipping point is shown for each functional group as a black dashed line and Table S1 gives additional information.](figures/Fig.S17_Medeiros_2025.png){#fig-medeiros-s17}

```{r tbl-medeiros-s1}
#| tbl-cap: "Medeiros et al. 2025, Table S1: change-point analysis for all 13 Lake Zurich functional groups. SS ratio: lower means stronger evidence of a shift (the 3 lowest are their clearest cases)."
# tribble() writes a small table row by row, which is easier to read than a
# column-wise data.frame(). The "~" marks the column names. These numbers are
# transcribed from Medeiros et al. 2025's published Table S1 -- they are their
# results, not ours, and serve as the reference we compare NLA-GP against.
table_s1 <- tribble(
  ~code, ~Group,                     ~`TP month`, ~`SS ratio`, ~Phosphate,
  "Cr2", "Cryptophytes (large)",      "1996-02",    0.940,       0.024,
  "Cr1", "Cryptophytes (small)",      "1988-10",    0.855,       0.043,
  "Cy",  "Cyanobacteria",             "1986-06",    0.991,       0.055,
  "Di2", "Diatoms (large)",           "2006-08",    0.976,       0.023,
  "Di1", "Diatoms (small)",           "1996-06",    0.961,       0.023,
  "Go2", "Gold algae (large)",        "1987-06",    0.961,       0.049,
  "Go1", "Gold algae (small)",        "1996-03",    0.989,       0.024,
  "Gr2", "Green algae (large)",       "1987-11",    0.853,       0.047,
  "Gr1", "Green algae (small)",       "1988-06",    0.996,       0.043,
  "Pr",  "Invertebrate predators",    "1989-10",    0.995,       0.046,
  "He",  "Large herbivores",          "1992-10",    0.980,       0.035,
  "Mi",  "Mixotrophic flagellates",   "1996-03",    0.931,       0.024,
  "Om",  "Omnivores",                 "1986-06",    0.903,       0.055
)
# Drop the short code column (we only need it in the code, not the display),
# sort by SS ratio so the strongest evidence sits at the top, and render.
table_s1 %>% select(-code) %>% arrange(`SS ratio`) %>% knitr::kable()
```

## Results

```{r run_zurich_all}
# Analysis settings, shared by all 13 groups:
E <- 3            # 3 lagged values as predictors
tau <- 1          # lags one month apart
Dskipstep <- 16   # try every 16th cut point (16 months) -- coarser = faster
bandwidth <- 32   # smoothing width for the error curve

# Library/test split, in months. Left check: library = the first 200 months,
# test = everything after. Right check: test = the first 60 months.
L_left <- 200; L_right <- 60
T_z <- nrow(z)    # total number of months in the series

# pmap() walks table_s1 row by row, handing each row's columns to the function
# below as named arguments -- so this runs the whole analysis once per group.
zurich_results <- pmap(table_s1, function(code, Group, `TP month`, `SS ratio`, Phosphate) {
  y <- z[[paste0("scaled_", code)]]   # this group's z-scored abundance series

  # Split Medeiros et al.'s date "1996-02" into year and month, and find the
  # matching row, so their answer is on the same index scale as ours.
  ym <- as.integer(strsplit(`TP month`, "-")[[1]])
  ref_idx <- zurich_month_index(z, ym[1], ym[2])

  # The two NLA checks: one looking for an early shift, one for a late shift.
  left <- nla_left(y, L = L_left, R = T_z, E = E, tau = tau, Dskipstep = Dskipstep, bandwidth = bandwidth)
  right <- nla_right(y, L = L_right, R = T_z, E = E, tau = tau, Dskipstep = Dskipstep, bandwidth = bandwidth)
  combined <- median(c(left$tau_hat, right$tau_hat), na.rm = TRUE)   # single answer
  tibble(code = code, Group = Group, `Medeiros date` = `TP month`, `SS ratio` = `SS ratio`,
         `NLA-GP left` = left$tau_hat, `NLA-GP right` = right$tau_hat,
         `NLA-GP combined` = combined,
         # Translate our row-index answer back into a calendar date, so it can
         # be read against Medeiros et al.'s date directly.
         `NLA-GP date` = if (is.na(combined)) NA_character_ else format(z$date[round(combined)], "%Y-%m"))
})
```

```{r tbl-zurich-summary}
#| tbl-cap: "NLA-GP detected change points vs. Medeiros et al. 2025 (Table S1), all 13 Lake Zurich functional groups, using the full 1977-2020 series and z-scored (not log-transformed) abundance."
# pmap() returned a list of one-row tables; bind_rows() stacks them into one.
zurich_summary <- bind_rows(zurich_results) %>% arrange(`SS ratio`)
knitr::kable(zurich_summary %>% select(-code))   # hide the short code column
```

```{r fig-zurich-nla-vs-medeiros}
#| fig-cap: "Lake Zurich functional groups (z-scored abundance) with Medeiros et al. 2025's change point (red) and NLA-GP's combined estimate (blue)."
#| fig-height: 12
# Reshape the 13 scaled columns into long form: one row per (date, group),
# which is the shape ggplot needs to draw one panel per group.
long_ts <- z %>%
  select(date, starts_with("scaled_")) %>%            # date + the 13 scaled columns
  pivot_longer(-date, names_to = "code", values_to = "value") %>%
  mutate(code = str_remove(code, "^scaled_")) %>%     # "scaled_Cy" -> "Cy"
  # left_join() attaches the readable group name by matching on the code.
  left_join(table_s1 %>% select(code, Group), by = "code")

# The two change-point dates to draw on each panel. Pasting "-01" onto
# "1996-02" makes a full date R can place on the time axis.
lines_df <- zurich_summary %>%
  mutate(medeiros_date = as.Date(paste0(`Medeiros date`, "-01")),
         # if_else() guards the groups where NLA-GP found nothing (NA).
         nla_date = if_else(is.na(`NLA-GP date`), as.Date(NA), as.Date(paste0(`NLA-GP date`, "-01"))))

ggplot(long_ts, aes(date, value)) +
  geom_line(linewidth = 0.3, color = "grey30") +      # the observed series
  # Two sets of vertical markers, drawn from lines_df rather than long_ts.
  # Naming the colour inside aes() is what makes the legend appear.
  geom_vline(data = lines_df, aes(xintercept = medeiros_date, color = "Medeiros"), linetype = "dashed") +
  geom_vline(data = lines_df %>% filter(!is.na(nla_date)),
             aes(xintercept = nla_date, color = "NLA-GP"), linetype = "dashed") +
  facet_wrap(~Group, ncol = 3, scales = "free_y") +   # one panel per group
  scale_color_manual(values = c(Medeiros = "firebrick", `NLA-GP` = "steelblue"), name = NULL) +
  labs(x = NULL, y = "z-scored abundance") +
  theme_minimal() +
  theme(legend.position = "top")
```

**We conclude:**

- Absolute timing differs from Medeiros et al.'s dates for most groups (@fig-zurich-nla-vs-medeiros) -- we would *not* claim precise agreement on the calendar date
- Medeiros et al.'s 3 clearest cases (lowest SS ratio: Green algae (large), Cryptophytes (small), Omnivores) are exactly the 3 groups we highlighted in earlier work; @tbl-zurich-summary shows how NLA-GP does on the full set of 13
- NLA-GP's relative ordering of these 3 clearest groups still tracks Medeiros et al.'s (Omnivores earliest, then Green algae (large), then Cryptophytes (small))

# PDO test

::: {.callout-tip title="Question"}
Huang et al. 2024 (Sec. 3.2) applied the original NLA-S-map to the Pacific Decadal Oscillation (PDO) index and detected a shift in its dynamics around 1973, using [monthly PDO data from NOAA](https://github.com/chromian/Nested-Library-Analysis/blob/main/Laboratory/data.csv) (as republished in their own analysis repository). Does NLA-GP find a similar shift, and does it agree with NLA-S-map?
:::

## Data

**To answer this, we:**

- Load the same monthly PDO series, and take a 12-month moving average, restricted to 1875-2000, exactly as Huang et al. 2024 describe
- Fix the **test set to 1875-1919** (the early part) and let the **library (1920-2000) shrink from the right** -- i.e. the right check, exactly as they did
- Compare NLA-GP and NLA-S-map's detected change point to their reported ~1973

```{r pdo_functions}
# skip = 1 ignores the file's first line, which is a header comment, not data.
pdo <- read.csv(here("data", "raw", "pdo.csv"), skip = 1)

# The Date column looks like "187501" (YYYYMM). substr() cuts out the pieces:
# characters 1-4 are the year, characters 5-6 the month.
pdo$year <- as.integer(substr(pdo$Date, 1, 4))
pdo$month <- as.integer(substr(pdo$Date, 5, 6))

pdo <- pdo[pdo$year >= 1875 & pdo$year <= 2000, ]   # window used by Huang et al.
pdo <- pdo[order(pdo$year, pdo$month), ]            # chronological order

# 12-month moving average: replace each month by the mean of it and the
# previous 11, which strips out the seasonal cycle. align = "right" means the
# window looks backwards; fill = NA leaves the first 11 months empty.
pdo$ma <- rollmean(pdo$Value, k = 12, fill = NA, align = "right")

pdo <- pdo[complete.cases(pdo), ]   # drop those first 11 empty months
pdo$idx <- seq_len(nrow(pdo))       # 1, 2, 3, ... = position in the series
```

```{r fig-pdo-traj}
#| fig-cap: "Monthly PDO index (grey) and its 12-month moving average (black), 1875-2000. Shaded: the test set (before 1920)."
# Row index of the first month of 1920: the boundary between test set and
# library. which.min(abs(...)) picks the row whose year is closest to 1920.
L_pdo <- which.min(abs(pdo$year - 1920))

# x = year expressed as a decimal, so months plot at the right position within
# the year (e.g. July 1950 = 1950.5).
ggplot(pdo, aes(year + (month - 1) / 12)) +
  # Grey band marking the test-set period. inherit.aes = FALSE stops this layer
  # from trying to use the x mapping above; -Inf/Inf mean "to the plot edge".
  geom_rect(xmin = -Inf, xmax = pdo$year[L_pdo], ymin = -Inf, ymax = Inf, fill = "grey90", inherit.aes = FALSE) +
  geom_line(aes(y = Value), color = "grey70", linewidth = 0.2) +   # raw monthly index
  geom_line(aes(y = ma), linewidth = 0.6) +                        # 12-month average
  labs(x = "Year", y = "PDO index") +
  theme_minimal()
```

## Results

We use a coarser `Dskipstep` than for Lake Zurich (24 months) since this series is much longer (1501 months); with it, each right check still takes a few minutes.

```{r run_pdo}
y_pdo <- pdo$ma        # analyse the smoothed (12-month average) series
T_pdo <- length(y_pdo)

# Only the RIGHT check here: the test set is the early period (pre-1920) and
# the library is trimmed from its late end -- exactly as Huang et al. did.
# Dskipstep = 24 means we step in 2-year jumps, since this series is long.
t0 <- Sys.time()   # start the stopwatch
right_pdo_gp <- nla_right(y_pdo, L = L_pdo, R = T_pdo, E = 3, tau = 1, Dskipstep = 24, bandwidth = 48)
time_pdo_gp <- as.numeric(Sys.time() - t0, units = "secs")   # elapsed seconds

# The same check with the original S-map forecaster, for comparison.
t0 <- Sys.time()
right_pdo_sm <- nla_right_smap(y_pdo, L = L_pdo, R = T_pdo, E = 3, tau = 1, Dskipstep = 24, bandwidth = 48)
time_pdo_sm <- as.numeric(Sys.time() - t0, units = "secs")
```

```{r tbl-pdo-summary}
#| tbl-cap: "NLA-GP vs. NLA-S-map on the PDO index (right check only, as in Huang et al. 2024), vs. their reported ~1973."
# Each method returns a row index; pdo$year[...] converts it into a calendar
# year. The is.na() guards handle the case where no change point was found.
tibble(Method = c("NLA-GP", "NLA-S-map"),
       `Detected year` = c(
         if (is.na(right_pdo_gp$tau_hat)) NA_character_ else as.character(round(pdo$year[round(right_pdo_gp$tau_hat)])),
         if (is.na(right_pdo_sm$tau_hat)) NA_character_ else as.character(round(pdo$year[round(right_pdo_sm$tau_hat)]))),
       `Huang et al. 2024` = c("~1973", "~1973"),
       `Runtime (s)` = round(c(time_pdo_gp, time_pdo_sm), 1)) %>%
  knitr::kable()
```

**We conclude:**

- NLA-GP lands closer to Huang et al.'s reported ~1973 than NLA-S-map does here -- the opposite of what we found on the synthetic food-chain benchmark, where S-map was far more accurate
- Taken together with the food-chain result, neither method is uniformly better: which one wins seems to depend on the system, not a general property of GP-EDM vs. S-map
- We did not reproduce Huang et al.'s original NLA-S-map result exactly (different smoothing bandwidth/step size, and their own implementation may differ from our reconstruction from S2 Text) -- treat this as a rough replication, not a precise one

# Conclusion

- On real Lake Zurich data (full series, all 13 functional groups, z-scored abundance): see @tbl-zurich-summary and @fig-zurich-nla-vs-medeiros for how NLA-GP compares to Medeiros et al. 2025's change-point analysis
- On the Pacific Decadal Oscillation index (Huang et al. 2024, Sec. 3.2): NLA-GP landed closer to their reported ~1973 shift than our reconstruction of NLA-S-map did (@tbl-pdo-summary) -- the reverse of the synthetic food-chain result in `1_implement.qmd`
- See `1_implement.qmd` for the algorithm and its validation on synthetic data with a known regime shift
